From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-30 Pi/paper/applications.tex
Work on section 7

Normalisation involves translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting back to
$\PiPlusLang$. This happends in several steps -- the evaluation step goes from $\PiLang$ to $\PiPlusLang$ reducing
$\times$ to $+$ and translating the corresponding combinators, then from $\PiPlusLang$ to $\PiHatLang$ reducing to
adjacent transpositions, then finally to $\UFin$ through words in $\Sn$ and Lehmer codes, and then, the quoting step
goes all the way back until $\PiPlusLang$. We define this $\normt$ function below using our previous evaluation
$\gdenot{\blank}$.

@@

@@

By normalisation, we also get a decision procedure for program equivalence -- two circuits are equivalent iff they
represent the same permutation, or equivalently, they have the same normal form. Recall the specification of reversible
disjunction from \cref{sec:examples}, the two Qiskit circuits for implementing it, and the corresponding $\PiLang$
definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$ normalises to
$\mathbb{8}+ \equiv \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + \mathbb{0})))))))$,
and normal forms for both circuits compute to the following, establishing their equivalence:

@@

Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is possible to
simply quote the desired permutation, synthesising a $\PiLang$ program. We use a function $\term{lookup}$ to write
tabulated permutations for readability.

@@

\noindent
The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
\textsf{true},\textsf{true}) is encoded as 011 or 3). The second entry maps index 1 (= 001) to the value 5 (= 101) which
states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Quoting this
permutation generates the same normalised program matching the desired structured types.

@@

More general, user-guided, reasoning can be done using the sound and complete level-2 combinators to rewrite $\PiLang$
programs.

>>>>

Normalisation happens by evaluating a combinator to a permutation and then reifying it back. The evaluation step goes
from $\PiLang$ to $\PiPlusLang$ reducing $\times$ to $+$ and translating the corresponding combinators, then from
$\PiPlusLang$ to $\PiHatLang$ reducing them to natural numbers and adjacent transpositions, then finally to $\UFin$
through words in $\Sn$ and Lehmer codes. Then, the quotation step goes all the way back until $\PiPlusLang$. We define
this composite $\normt$ function below using our previous evaluation $\gdenot{\blank}$.

@@

As an example, consider the \Afun{reversibleOr1} circuit introduced in~\cref{sec:pi}. After applying $\normt_{1}$, the
type $\mathbb{B}\ 3$ normalises to $\mathbb{8}+ \equiv \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
  (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + \mathbb{0})))))))$, and the circuit reduces to the following
normal form:

@@

By normalisation, we also get a decision procedure for program equivalence -- two circuits are equivalent iff they
represent the same permutation, or equivalently, they have the same normal form. Recall that in~\cref{sec:examples} we
introduced two different ways of computing reversible disjunction using two Qiskit circuit implementations, and then
in~\cref{sec:pi} showed the corresponding $\PiLang$ definitions: \Afun{reversibleOr1} and \Afun{reversibleOr2}.

After applying $\normt_{1}$ to each circuit, we verify that they both reduce to the same normal form
\Afun{reversibleOrNorm}, along with a 2-combinator witnessing their equivalence with the normal form -- by composing
these, we get the 2-combinator witnessing their equivalence.

@@

Instead of manually producing $\PiLang$-programs to implement the reversible disjunction specification, it is possible
to simply write down the permutation directly and then quote it, synthesising a $\PiLang$ program. We use a function
$\term{lookup}$ to write tabulated permutations for readability.

@@

\noindent
The permutation uses the canonical encoding of sequences of bits as natural numbers, e.g., {(\textsf{false},
\textsf{true},\textsf{true})} is encoded as 011 or 3. The second entry in the $\term{lookup}$ table maps index 1 (= 001)
to the value 5 (= 101), following the reversible disjunction specification (recall that
$\mathsf{reversibleOr}(h,b_1,b_2) = (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
and~$\underline{\vee}$ is exclusive-disjunction). Quoting this permutation generates the same normalised program
\Afun{reversibleOrNorm}, matching the desired structured type of a 1-combinator on $\mathbb{8}+$.

@@

In our development, we described a number of different representations of permutations: $\PiLang$ programs, lists of
transpositions, Lehmer codes, and automorphisms of finite sets. The equivalences between these representations allows us
to transfer theorems about permutations, proving them on the representations most suited for the particular task and
getting others for free.

As an example, consider the following problem. Toffoli gate on 3 bits can be extended to 4 bits, inserting a wire in
four possible positions. The function of such gate is then identity on the newly inserted bit, and controlled-controlled
not on the three remaining ones.

\medskip
\extendedToffoli3{}

On the other hand, it is possible to construct a proper 4-bit toffoli gate:

\medskip
\toffoli4{}

Now, the question we want to asnwer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the
behaviour of a 4-bit toffoli gate?

Solving this problem using the circuit representation is difficult. However, we can observe that 3-bit toffoli gate
implements an even permutation, that is, a permutation consisting of an even number of transpositions. On the other
hand, 4-bit toffoli gate is an odd permutation. It is an easily proven lemma that composition of even permutations
always produces an even permutation, which answers the question to the negative.

In our case, we define the $\term{parity}$ function:
parity : {n m : ℕ} → (c : S n Pi^.⟷₁^ S m) → ⊤ ⊔ ⊤
parity {X} {Y} c = list-len-parity (pi^2list c)

Thus, we formulate the theorem using the representation on lists of transpositions, and transfer it to circuits using
the equivalences proven in the previous sections.

% More general, user-guided, reasoning can be done using the sound and complete level-2 combinators to rewrite $\PiLang$
% programs.

### 2021-10-30 Pi/paper/examples.tex
Work on section 7

For concreteness, we demonstrate two different circuits that implement the following reversible function specification
$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface \verb|qasm|
and the graphical interface, in~\cref{fig:qiskit}.

>>>>

For concreteness, we demonstrate two different circuits that implement the following reversible function specification
$\mathsf{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface \verb|qasm|
and the graphical interface, in~\cref{fig:qiskit}.
